Tuning parameter selection for a penalized estimator of species richness

### 2019-07-31 rre_manuscript/paper.tex
edit sec 3, 4, 5.1

\amy{Happy to discuss, but I think it's probably easiest for me to be corresponding author, because you won't have your UW e-mail forever}
\alex{Having you as the corresponding author sounds great, just didn't want to volunteer you for unwanted emails until we discussed!}

@@

In this paper we consider the penalized maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter. Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross validation) cannot be employed. We propose using biological replicates to aid tuning.%, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.

@@

The classical model for estimating species abundance is a Poisson mixture model.  Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review). While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, sal first considered by \citet{fisher_1943}.

@@

Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$, and write $\eta = (\alpha, \delta)$.
% Then the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
% \begin{align}
% p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x.
% \end{align}
% Note that we are interested in estimating the parameters of a negative binomial
% a zero-truncated negative binomial
%
There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model. We will focus on frequentist maximum likelihood in this paper.

@@

The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.  A complication is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, and so derivative-based optimization methods are not appropriate.  This issue has been studied by \citet{lindsay_1987}, who provide a discrete analog to the score function for this model.  %However, they point out that the asymptotics required use $C \to \infty$, which may not be appropriate for all problems.
Two other approaches to likelihood maximization are the \textit{conditional} and \textit{profile} approaches.  The conditional approach \citep{sanathanan_1977} involves writing the likelihood as the product
\begin{align}
 L(C, \eta) &= L_b(C, \eta)L_c(\eta) \label{eq:likelihood}
\end{align}
where
 \begin{align}
 L_b(C, \eta) &= \frac{C!}{(C-c)! \ c!} \left[1 - p_{\eta}(0) \right]^c \left[ p_{\eta}(0) \right]^{C-c}  %& \text{(``Binomial'' component)}
  \label{eq:binomial_likelihood}\\
 L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k}. \label{eq:conditional_likelihood} % & \text{(Multinomial component/``conditional'')}
\end{align}
We see that the likelihood is the product of a binomial probability mass function $c \sim$ Binomial $\left(C, 1- p_{\eta}(0)\right)$, and a multinomial probability mass function $(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$.
This decomposition is convenient because $L_c$ is a function of $\eta$ alone.
% $L_b$ shows the probability of observing $c$ species follows a binomial distribution, i.e. $c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$.  Letting $\kmax$ be the largest $k$ such that $f_k > 0$, $L_c$ shows that observing the frequency counts given by $f_1, f_2, \dots , f_{\kmax}$ follows a multinomial distribution given $c$.  The multinomial cell probabilities are proportional to $p_{\eta}(k)$.
\citet{sanathanan_1977}'s conditional approach to optimization involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
The conditional estimate of $C$ is then
\begin{equation}
\widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor \label{eq:conditional_c}
\end{equation}
where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.  In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).  The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
\amy{Need to justify why we are bringing this up}

% \citet{sanathanan_1977} uses this decomposition to propose the conditional approach to likelihood maximization, which is done in two steps (see \citet{chao_2002} for a detailed presentation).  First we maximize $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$.  Then $L_b(C, \widehat{\eta}_c)$ is maximized over $C$.  With $\eta$ fixed there is a closed form expression for this second step,
% \begin{equation}
% \widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor %\label{eq:conditional_c}
% \end{equation}
% where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.  In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).  The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.

@@

A perennial issue in the species problem, especially for microbial datasets, is the instability of species richness estimators \citep{Rocchetti:2011tv,willis_2015}.
%
% Under the gamma-Poisson model,
%
% To illustrate, consider that a first-order approximation gives $$Var(\hat{C}_c | c) \approx \frac{c^2}{(1-\hat{p}(0))^4}Var(\hat{p}(0))$$ \amy{check this}
% Therefore,
% , especially maximum likelihood estimators.  The estimates often have unacceptably high variance and can become dramatically inflated.
%
%\sout{\textcolor{blue}{Since $Pr(X = 0) = \left( \frac{\delta}{1 + \delta}\right)^\alpha$, if there are many low abundance species then $\alpha$ will be near 0. Many high abundance species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.  The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.}}
%% One way of improving stability is to abandon the Poisson mixture paradigm entirely.  Though we will focus on the gamma-Poisson model in this paper, we would be remiss if we did not briefly mention other popular alternatives.  Coverage-based estimators, which use the proportion of species observed $k$ times rather than the actual counts were originally proposed by \citet{good_1953} and remain popular.  Capture-recapture methods (see \citet{chao_1987}) have a deep literature in animal abundance studies but are not as applicable in microbial data where the sampling  destroys the organism.  See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.
A proposal which encourages stability of estimates under the gamma-Poisson model is due to \citet{wang_2005}.  Their proposal is to add a penalty term to the log likelihood that penalizes the probability of observing a species zero times. They consider the
penalized log likelihood
\begin{equation}
\ell_\lambda(C, \eta) = \ell(C,\eta) - \lambda \log p_{\eta}(0)
\label{eq:wang_lindsay}
\end{equation}
where $\ell(C,\eta) = \log L(C, \eta)$ is the log likelihood and $\lambda > 0$ is a penalization parameter. For $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda < \hat{C}_{MLE}$
\amy{up to here with editing}
for $\eta$ to fit so that $p_{\widehat{\eta}}(0)$ is smaller. \amy{Check this: Therefore $\alpha$ and $\delta$ are further from the boundary, which alleviates some of the instability.} The penalization term also causes a monotonic decrease in $C$ estimates as greater values of $\lambda$ are used. For a large enough $\lambda$, $\hat{C}$ will shrink to the observed richness $c$.

\citet{wang_2005} also consider two other penalty functions, which we do not discuss here.

In their paper they show a trade-off exists: greater values of $\lambda$ correspond to a more stable estimator, but at the potential cost of negative bias.  This is a realization of the bias-variance tradeoff.
The choice of penalization parameter implies a preference for lower variance or lower bias, and therefore adds a degree of subjectivity to the estimation procedure (comparable to a Bayesian prior as discussed in \citet{wang_2005}).  We would also expect different data sets to require different $\lambda$ values for optimal estimation, which makes direct application of this method to real data challenging.  Our primary interest in studying the analysis of replicate frequency count tables is to leverage the replicates to tune the penalization parameter in Equation \ref{eq:wang_lindsay}.  The goal of this exploration is to obtain an estimator which is more stable than the unpenalized MLE but does not have the disadvantage of requiring a fixed, user-specified penalization parameter.

@@

Throughout we will focus on the gamma-Poisson model with penalization as proposed by \citet{wang_2005}.  In this section we extend the penalized likelihood for a single frequency count table to the case where multiple frequency count tables have been collected.

Let $r$ be the number of frequency count tables.  We will assume that these $r$ frequency count tables are biological replicates drawn independently from the same population, i.e. they have a common structure specified by the parameters $C$ and $\eta$.  A set of $r$ frequency count tables from the same population will be called a \textit{sample}, and a single frequency count table will be referred to as a \textit{replicate}.  Let the number of species observed $k$ times in replicate $j$ be $f_{kj}$, $j \in \{1, \dots , r\}$.  Let $c_j$ be the observed richness in replicate $j$.

Microbial data is often available as labeled abundance information $\{ X_{ij} \}$, the number of times species $i$ was observed in replicate $j$.  In this paper we study data represented in frequency counts $\{f_{kj}\}$ and it is worth pointing out that we lose some information in this representation.  For instance, in sample of frequency count tables it cannot be determined whether species $i$ was observed in two different replicates.  By using a mixed Poisson model we assume a common abundance structure between the species in order to estimate $C$.  While there may be other methods which leverage the additional information in an abundance table representation, the frequency counts are a minimal sufficient statistic for mixed Poisson models.

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.  Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.  We will refer to this as the \textit{penalized log likelihood} and it is given by
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) =& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate using $\lambda$, that is
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.

We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
\end{equation}

To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.  While it may be faster and asymptotically equivalent to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.  In this paper our end goal is to explore tuning, so the stability and accuracy of our underlying optimization algorithm have priority over speed.  Crucially, will use exactly the same optimization for the likelihood throughout all simulations in this paper.

To implement a direct optimization approach we search over candidate $C$ values in a grid, respecting the fact that $C$ is a discrete parameter.  A gradient-based search over $\eta$ values is used at each $C$ in the grid.  The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.  While this paper does not focus on optimization, our code is made available to the reader (see Section \ref{sec:code} for links).

@@

Establishing that penalization can improve richness estimates is a crucial foundation before we attempt to evaluate tuning methods.  \citet{wang_2005} established that it can improve estimates for $r = 1$ and in their particular application area.  We need to show this for our models and $r > 1$.  We propose to do so by simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.  If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) over all simulations is less than RMSE($\widehat{C}_0)$ we conclude that penalization is effective under those simulation parameters.

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates.  The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.

In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.  We make this information available here for the interested reader but delay our discussion of the differences until Section \ref{sec:tuning_simulations} where they strongly influence our results.

>>>>

@@

In this paper we consider the penalized maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter. Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross-validation) cannot be employed. We propose using biological replicates to aid tuning. %, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.

@@

The classical model for estimating species abundance is a Poisson mixture model.  Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review). While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{fisher_1943}.

@@

Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$, and write $\eta = (\alpha, \delta)$.
% Then the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
% \begin{align}
% p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x.
% \end{align}
% Note that we are interested in estimating the parameters of a negative binomial
% a zero-truncated negative binomial
%
There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model. We will focus on frequentist maximum likelihood in this paper.
\amy{introduce Gamma later?}

@@

The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.  A complication is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, and so derivative-based optimization methods are not appropriate.  This issue has been studied by \citet{lindsay_1987}, who provide a discrete analog to the score function for this model.  %However, they point out that the asymptotics required use $C \to \infty$, which may not be appropriate for all problems.
Two other approaches to likelihood maximization are the \textit{conditional} and \textit{profile} approaches.  The conditional approach \citep{sanathanan_1977} involves writing the likelihood as the product
\begin{align}
 L(C, \eta) &= L_b(C, \eta)L_c(\eta) \label{eq:likelihood}
\end{align}
where
 \begin{align}
 L_b(C, \eta) &= \frac{C!}{(C-c)! \ c!} \left[1 - p_{\eta}(0) \right]^c \left[ p_{\eta}(0) \right]^{C-c}  %& \text{(``Binomial'' component)}
  \label{eq:binomial_likelihood}\\
 L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k}. \label{eq:conditional_likelihood} % & \text{(Multinomial component/``conditional'')}
\end{align}
We see that the likelihood is the product of a binomial probability mass function $c \sim$ Binomial $\left(C, 1- p_{\eta}(0)\right)$, and a multinomial probability mass function $(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$.
This decomposition is convenient because $L_c$ is a function of $\eta$ alone.
% $L_b$ shows the probability of observing $c$ species follows a binomial distribution, i.e. $c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$.  Letting $\kmax$ be the largest $k$ such that $f_k > 0$, $L_c$ shows that observing the frequency counts given by $f_1, f_2, \dots , f_{\kmax}$ follows a multinomial distribution given $c$.  The multinomial cell probabilities are proportional to $p_{\eta}(k)$.
\citet{sanathanan_1977}'s conditional approach to optimization involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
The conditional estimate of $C$ is then
\begin{equation}
\widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor \label{eq:conditional_c}
\end{equation}
where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.  In the \textit{profile} likelihood approach the expression in Equation \eqref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).  The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
\amy{Need to justify why we are bringing this up}

% \citet{sanathanan_1977} uses this decomposition to propose the conditional approach to likelihood maximization, which is done in two steps (see \citet{chao_2002} for a detailed presentation).  First we maximize $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$.  Then $L_b(C, \widehat{\eta}_c)$ is maximized over $C$.  With $\eta$ fixed there is a closed form expression for this second step,
% \begin{equation}
% \widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor %\label{eq:conditional_c}
% \end{equation}
% where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.  In the \textit{profile} likelihood approach the expression in Equation \eqref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).  The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.

@@

A perennial issue in the species problem, especially for microbial datasets, is the instability of species richness estimators \citep{Rocchetti:2011tv,willis_2015}.
%
% Under the gamma-Poisson model,
%
% To illustrate, consider that a first-order approximation gives $$Var(\hat{C}_c | c) \approx \frac{c^2}{(1-\hat{p}(0))^4}Var(\hat{p}(0))$$ \amy{check this}
% Therefore,
% , especially maximum likelihood estimators.  The estimates often have unacceptably high variance and can become dramatically inflated.
%
%\sout{\textcolor{blue}{Since $Pr(X = 0) = \left( \frac{\delta}{1 + \delta}\right)^\alpha$, if there are many low abundance species then $\alpha$ will be near 0. Many high abundance species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.  The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.}}
%% One way of improving stability is to abandon the Poisson mixture paradigm entirely.  Though we will focus on the gamma-Poisson model in this paper, we would be remiss if we did not briefly mention other popular alternatives.  Coverage-based estimators, which use the proportion of species observed $k$ times rather than the actual counts were originally proposed by \citet{good_1953} and remain popular.  Capture-recapture methods (see \citet{chao_1987}) have a deep literature in animal abundance studies but are not as applicable in microbial data where the sampling  destroys the organism.  See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.
A proposal which encourages stability of estimates under Poisson-mixture models is due to \citet{wang_2005}.  Their proposal is to add a penalty term to the log likelihood that penalizes the probability of observing a species zero times. They consider the
penalized log likelihood
\begin{equation}
\ell_\lambda(C, \eta) = \ell(C,\eta) - \lambda \log p_{\eta}(0)
\label{eq:wang_lindsay}
\end{equation}
where $\ell(C,\eta) = \log L(C, \eta)$ is the log likelihood and $\lambda > 0$ is a penalization parameter.
Let $\hat{C}_\lambda = \arg \max_{C} \ell_\lambda(C, \eta)$.
Then, for $\lambda \geq \lambda',$ $\hat{C}_\lambda \leq \hat{C}_{\lambda'}$ \citep[Theorem 1]{wang_2005}. In particular, for $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda$ is less than or equal to $\hat{C}_0$, the maximum likelihood estimate. Furthermore, for a large enough $\lambda$, $\hat{C}_\lambda = c$; that is, the penalised maximum likelihood estimate shrinks to the observed richness $c$.
\citet{wang_2005} also consider two other penalty functions of similar form, which we do not discuss here except to note that our tuning parameter selection methods would equally apply.

\citet{wang_2005} show that a trade-off exists: greater values of $\lambda$ correspond to a more stable estimator, but at the potential cost of negative bias.
The choice of penalization parameter implies a preference for lower variance or lower bias, adding subjectivity to the estimation procedure.
\citet{wang_2005} note that a limitation of their proposal is that ``one must select a penalty function and a tuning parameter, and it is nigh impossible to make convincing statements about why one choice should be uniformly superior to another.''
Furthermore, we expect different data sets to require different $\lambda$ values for optimal mean squared error.
% Our primary interest in studying the analysis of replicate frequency count tables is to leverage the replicates to tune the penalization parameter in Equation \eqref{eq:wang_lindsay}.
The goal of this paper is to propose and investigate data-adaptive methods to select $\lambda$.

@@

We focus on the gamma-Poisson model and penalised log-likelihood of the form \eqref{eq:wang_lindsay}, and consider the case where we observe $r$ independent frequency count tables from the population under study.
% with penalization as proposed by \citet{wang_2005}.  In this section we extend the penalized likelihood for a single frequency count table to the case where multiple frequency count tables have been collected.
We will assume that these $r$ frequency count tables are biological replicates drawn independently from the same population, i.e. they have a common structure specified by the parameters $C$ and $\eta$.  A set of $r$ frequency count tables from the same population will be called a \textit{sample}, and a single frequency count table will be referred to as a \textit{replicate}.  Let the number of species observed $k$ times in replicate $j$ be $f_{kj}$, $j \in \{1, \dots , r\}$.  Let $c_j$ be the observed richness in replicate $j$.

% Microbial data is often available as labeled abundance information $\{ X_{ij} \}$, the number of times species $i$ was observed in replicate $j$.  In this paper we study data represented in frequency counts $\{f_{kj}\}$ and it is worth pointing out that we lose some information in this representation.  For instance, in sample of frequency count tables it cannot be determined whether species $i$ was observed in two different replicates.  By using a mixed Poisson model we assume a common abundance structure between the species in order to estimate $C$.  While there may be other methods which leverage the additional information in an abundance table representation, the frequency counts are a minimal sufficient statistic for mixed Poisson models.

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.
Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.
We define the \textit{penalized log likelihood} for multiple samples as
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) :=& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate for based on tuning parameter $\lambda$, that is,
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.

We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
\end{equation}

To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.  While it may be faster to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.
Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed.

To implement a direct optimization approach we search over candidate $C$ values in a grid.  A gradient-based search over $\eta$ values is used at each $C$ in the grid to find the maximum penalised likelihood  $(C, \eta)$.

@@

Before evaluating potential tuning parameter selection methods, we first establish that penalization can improve richness estimation.  \citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.
We propose simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.  If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective for the given data generating mechanism.

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$.  We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017}.
For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
% To select $\eta$, we investigated the unregularised
%  two microbial datasets
% so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.
In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see. We return to a discussion of the different data structures in Section \ref{sec:tuning_simulations}.
For each combination of $\eta$ and $r$, we performed 100 simulations.
We investigated the grid $\lambda \in \{0, 5, \dots , 120\}$, and found that this grid was sufficiently expansive to ensure that the RMSE-minimising $\lambda$ was not on the boundary of the grid
(see Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}).

% One final setting is necessary for these simulations: a set of $\lambda$ values to test over.  Our primary concern is to ensure the set covers the optimal value, i.e. $\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.  While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to confirm that the set was sufficiently large after the simulation is complete.  As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.  If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ $\widehat{C}_{\lambda}$ will be further from $C$, and the RMSE will increase.  Then we can be assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.
% This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}. Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\tblcaption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.  We also give the expected frequency count of the most abundant species when $C = 1000$.
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\
   \hline
\end{tabular}
\normalsize
\end{table}
